\chapter{Operator-Theoretic Learning: Koopman, DMD, SINDy}
\label{ch:operator-theoretic}

\epigraph{\itshape Linearity is a virtue. Even nonlinear systems are
linear --- in infinite dimensions.}{Bernard Koopman, 1931}

\section{The Koopman operator}
For a discrete dynamical system $\bm{x}_{k+1}=\bm{F}(\bm{x}_k)$ on a
state space $M$, the \emph{Koopman operator} $\Kop$ acts on observables
$g:M\to\C$:
\begin{equation}
(\Kop g)(\bm{x}) = g(\bm{F}(\bm{x})).
\end{equation}
$\Kop$ is linear (even when $\bm{F}$ is nonlinear) but
infinite-dimensional. Its spectrum encodes invariant sets, modes, and
ergodic properties.

\subsection{Dynamic Mode Decomposition (DMD)}
Given snapshots $\bm{X}=[\bm{x}_0,\dots,\bm{x}_{m-1}]$,
$\bm{Y}=[\bm{x}_1,\dots,\bm{x}_m]$, DMD seeks $\bm{A}$ minimising
$\|\bm{Y}-\bm{A}\bm{X}\|_F$, giving $\bm{A}=\bm{Y}\bm{X}^\dagger$.
The eigendecomposition of $\bm{A}$ yields DMD modes (spatial structures)
and DMD eigenvalues (frequencies and growth rates).

\subsection{Extended DMD (eDMD)}
Lift state via a dictionary
$\bm{\psi}(\bm{x})=(\psi_1(\bm{x}),\dots,\psi_N(\bm{x}))$ and
regress $\bm{\psi}(\bm{x}_{k+1})\approx \bm{K}\bm{\psi}(\bm{x}_k)$.
For polynomial dictionaries this recovers SINDy; for radial bases /
random features this recovers kernel DMD.

\subsection{Deep Koopman}
Replace the dictionary by a learned encoder
$\bm{\psi}_\theta:M\to\C^N$ with a \emph{linear} latent dynamics
$\bm{z}_{k+1}=\bm{K}_\theta\bm{z}_k$ and a decoder
$\bm{\phi}_\theta$. Loss combines reconstruction, prediction, and
linearity terms~\cite{lusch2018koopman}.

\begin{insight}{Why Koopman is powerful for ML}
\begin{itemize}
\item Eigenfunctions provide globally meaningful coordinates --
``intrinsic'' phase variables.
\item Linearity in latent space yields easy long-horizon prediction,
control, and stability analysis.
\item Captures multi-scale temporal structure via Koopman spectra.
\end{itemize}
But \emph{spectral pollution} --- spurious eigenvalues from finite
truncation --- is a persistent danger. Modern remedies include
ResDMD, EDMD-DL with closure error, and residual-based residual
estimators.
\end{insight}

\section{Sparse identification of nonlinear dynamics (SINDy)}
SINDy~\cite{brunton2016sindy} fits the ODE
$\dot{\bm{x}}=\bm{f}(\bm{x})$ in a sparse linear combination of a
candidate library $\bm{\Theta}(\bm{x})$:
\begin{equation}
\dot{\bm{X}} = \bm{\Theta}(\bm{X})\bm{\Xi},\qquad
\bm{\Xi}=\argmin_{\bm{\Xi}}\|\dot{\bm{X}}-\bm{\Theta}(\bm{X})\bm{\Xi}\|_2^2 + \lambda\|\bm{\Xi}\|_0.
\end{equation}
$\ell_0$ is approximated by sequential thresholding ($\text{STLS}$),
LASSO, or sparse relaxed regularised regression (SR3).

\subsection{Variants}
\paragraph{Weak-form / Integral SINDy.}
Multiply the ODE by test functions and integrate by parts to avoid
finite-difference noise on $\dot{\bm{x}}$. Weak-SINDy and PDE-FIND for
PDEs.
\paragraph{Constrained SINDy.}
Enforce conservation, energy, or symmetry by linear constraints on
$\bm{\Xi}$.
\paragraph{Ensemble SINDy / E-SINDy.}
Bootstrap the data and library to obtain robust models with uncertainty
estimates.
\paragraph{Implicit SINDy.}
Identify rational models $f=\bm{\Theta}_n\bm{\xi}_n / \bm{\Theta}_d\bm{\xi}_d$.
\paragraph{SINDy-AE.}
Wrap SINDy in an autoencoder to discover dynamics in a learned
coordinate system.

\subsection{Pitfalls}
\begin{pitfall}{Library construction is the whole game}
SINDy is interpretable only if the true dynamics live in your library.
Choosing too small a library forces wrong models; too large makes
sparsity selection brittle. Domain knowledge wins.
\end{pitfall}

\section{Symbolic regression}
Where SINDy fixes a candidate library, symbolic regression searches the
space of expression trees (operators, variables, constants):
\begin{itemize}
\item \textbf{Genetic programming}: PySR, Eureqa.
\item \textbf{AI Feynman}: combines neural fits, dimensional analysis,
and combinatorial search to recover Feynman lectures' equations.
\item \textbf{Transformer-based symbolic regression}: SymbolicGPT,
NeSymRes, end-to-end equation generation.
\item \textbf{KAN-SR}: Kolmogorov--Arnold networks (\cref{ch:hybrid-ude})
yield analytic primitives via spline activations.
\end{itemize}
Symbolic regression provides interpretable physics --- closing the loop
from data to equations.

\section{Reservoir computing and echo state networks}
Random nonlinear recurrent reservoirs followed by a linear readout fit
chaotic dynamics surprisingly well, with $\Oh(\text{seconds})$ training
time. Particularly good for short-term forecasting of high-dimensional
spatiotemporal chaos (Lorenz-96, Kuramoto--Sivashinsky), and as a
baseline against deep models.

\section{Recipe: pick the right operator-theoretic tool}
\begin{recipe}{Decision flow}
\begin{enumerate}
\item Have snapshots of a single trajectory and want modes? \emph{DMD}.
\item Want a global linear model on lifted observables? \emph{eDMD /
deep Koopman}.
\item Suspect the dynamics are sparse polynomials in known variables?
\emph{SINDy}.
\item No idea of the right form, want a closed-form expression?
\emph{symbolic regression}.
\item Just want a fast forecaster on a chaotic system? \emph{reservoir
computing} as baseline.
\end{enumerate}
\end{recipe}

\section{Connections to neural operators}
Koopman is, abstractly, an \emph{operator on observables}; FNO and
DeepONet are \emph{operators on PDE solution states}. Both reformulate
nonlinear dynamics as linear/operator problems; the difference is the
ambient space (point observables vs.\ function fields). Several works
connect the two perspectives, e.g.\ Koopman neural operators and
spectral DMD-FNO hybrids.
